---
title: Comparison of data manipulation with different R and Python packages Part I
author: Omer Acar
date: '2019-08-05'
slug: comparison-of-data-packages
draft: false
categories:
  - R
  - Python
tags:
  - dplyr
  - data.table
  - pandas
  - R
  - python
subtitle: ''
summary: ''
authors: []
lastmod: '2019-08-05T14:41:37-04:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---


I have been using tidyverse functions with dplyr or tidyr for data manipulation mainly. However there are times where I had to use Python for due to need for a specific package or collaboration with people using only Python, thus needed to use Pandas for similar purposes. However I am in no way a Python expert and everytime I needed to look up a lot of basic things and end up just going back and forth between Python and R. Here I wanted to create a reference page with examples for myself and people who are interested in same things to use Pandas and Tidyverse functions for data manipulation purposes. 


```{r, include=FALSE}
knitr::opts_chunk$set(message=FALSE)
```

First to use python in R markdown, I use RStudio and reticulate
```{r}
library(reticulate)
use_condaenv("r-reticulate")
```

Then let's load the libraries and get the data, I'll use iris dataset that's available by default in R:

### R

```{r}
library(dplyr)
library(data.table)
data("iris")
#iris
```

### Python
You can get the same data from seaborn-data link.

```{python}
import pandas as pd
iris = pd.read_csv('https://raw.githubusercontent.com/mwaskom/seaborn-data/master/iris.csv')
iris
```

Tidyverse functions are usually self-explained. To get rows with a condition there is a ```filter``` function, to extract a column from data frame there is ```select```. ```mutate``` is used for adding new columns to dataframe based on values on other columns or as I use most of time with external ```ifelse``` statements. Let's see how these functions and uses are in both languages.
To actually compare functions and to not confuse people who are not familiar with pipes in tidyverse, I will omit using pipes and use dataframes as arguments to functions.


## Filter
Let's take the rows with Sepal Length > 5 condition.

### R
```{r}
iris_sepal_long <- filter(iris, Sepal.Length>5)
head(iris_sepal_long)
```

### Python
```{python}
iris_sepal_long = iris[iris.sepal_length>5]
iris_sepal_long
```


Any logical statements combined with logical `OR` or `AND` signs can be used when filtering data frame. I'll show with python but same applies to R as well. For example we can select rows with Sepal length > 5 and Sepal width < 4 in `setosa` species in the following way:

```{python}
iris_sepal_filter = iris[(iris.sepal_length>5)&(iris.sepal_width<4)&(iris.species=='setosa')]
iris_sepal_filter
```


## Mutate

Let's make another data frame that has a column of Sepal Length to Sepal Width ratio. While ```mutate``` is the function in R, ```assign``` will do the same thing in Python.

### R

```{r}
#iris$newcolumn <- iris$Sepal.Length/iris$Sepal.Width
iris_sepal_length_over_width <- mutate(iris,sepal_length_over_width = Sepal.Length / Sepal.Width)
head(iris_sepal_length_over_width)
```


### Python

```{python}
iris_sepal_length_over_width = iris.assign(sepal_length_over_width = iris.sepal_length / iris.sepal_width)
iris_sepal_length_over_width
```


## Select
This is one of the most important functions when dealing with data frames. You don't always need all the data in a data frame but only some columns.

Let's select columns related with Sepal data.

### R

```{r}
sepal_data <- select(iris,Sepal.Length,Sepal.Width)
head(sepal_data)
```


### Python

```{python}
sepal_data = iris[['sepal_length','sepal_width']]
sepal_data
```

However one thing I do a lot when using ```select``` getting a column and turn it into a vector because `select` returns a data frame with a signle column, for quick plotting with base R functions. To do this in R you need to use `pull` function after selecting the data. Then you will have a vector of the selected column which can be directly used in plotting functions like `plot` or `hist`

### R

```{r}
sepal_width_vector <- pull(select(iris,Sepal.Width))
head(sepal_width_vector)
hist(sepal_width_vector)
```

### Python

I think there would be different solution to this in Python way but to directly replicate what I did above in R you can call tolist() function on the column data.

```{python}
sepal_width_vec = iris.sepal_width.tolist()
sepal_width_vec
import seaborn as sns

sns.set_style('darkgrid')
sns.distplot(sepal_width_vec)
```


## Conclusions

Here I compared the mostly used functions in data science when dealing with data frames in R using dplyr functions and in Python using Pandas. These are only single ways of doing these things and R has multiple different ways of doing same things due to difference in base R data frame functions and packages like `dplyr` or `data.table`. I feel `tidyverse` way of doing things are intuitive and easy to read especially using pipes. 

As far as I know Pandas is the mostly used package for working with data frames in Python. I am sure there are other packages but I never need to work with them or encountered. Pandas website also has a [reference page](https://pandas.pydata.org/pandas-docs/stable/getting_started/comparison/comparison_with_r.html) for R and Pandas function comparisons.



<!-- ## Sorting -->
<!-- ```{r} -->
<!-- arrange(iris) -->
<!-- ``` -->

<!-- ## Summarizing data with groups -->